---
title: "amcmc: A flexible MCMC estimation framework"
output: github_document
---

[![Travis-CI Build Status](https://travis-ci.org/USCbiostats/amcmc.svg?branch=master)](https://travis-ci.org/USCbiostats/amcmc)
[![Build status](https://ci.appveyor.com/api/projects/status/3x9qj7imvoijb1vf?svg=true)](https://ci.appveyor.com/project/gvegayon/amcmc)
[![Coverage Status](https://img.shields.io/codecov/c/github/USCbiostats/amcmc/master.svg)](https://codecov.io/github/USCbiostats/amcmc?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/amcmc)](https://cran.r-project.org/package=amcmc)


Current features:

1.  Automatic stop using convergence checker.

2.  Parallel chains using `parallel::lapply`

3.  Normal Random Walk with Reflective Boundaries kernel.


# Installing

From github:

```r
devtools::install_github("USCbiostats/amcmc")
```

# Example

Linear regression model

```{r}
library(amcmc)

# Simulating data
set.seed(78845)
n <- 1000
X <- rnorm(n)
y <- 3 + 2*X + rnorm(n, sd = 2)

# Loglikelihood function
ll <- function(par, X., y.) {
  
  ans <- sum(log(dnorm((y. - (par[1] + X.*par[2]))/par[3])/par[3]))
  
  if (!is.finite(ans))
    return(-Inf)
  
  ans
}

# Running the MCMC
ans <- MCMC(
  ll, X. = X, y. = y,
  initial = c(1, 1, 1),
  nsteps   = 2e4,
  nchains  = 4L,
  autostop = 1e3,
  burnin   = 1e4,
  scale    = .1,
  multicore = TRUE
  )
```

```{r summary-and-plot1}
library(coda)

summary(ans)
plot(ans)
gelman.diag(ans)
```


# Other tools

*   https://cran.r-project.org/web/packages/mcmc/mcmc.pdf

*   https://cran.r-project.org/web/packages/HybridMC/HybridMC.pdf

*   https://cran.r-project.org/web/packages/adaptMCMC/adaptMCMC.pdf

*   https://cran.r-project.org/web/packages/elhmc/elhmc.pdf

# Contributing to `amcmc`

Please note that the 'amcmc' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

# Funding

Supported by National Cancer Institute Grant #1P01CA196596.