---
title: "Hierarchical Bayes Model"
author: "George Vega Yon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Suppose that you have a dataset in which the observed process follows a normal
distribution in the form of $N(\mu_c, 1)$, this is, each observation follows
a normal distribution with paramater $(\mu_c, 1)$, where $\mu_c \sim N(\gamma, 1)$
and $\gamma \sim \mbox{Unif}(-1,1)$. So for each class of observation, the mean
was drawn from a normal distribution with unknown mean, and we assume the mean
is within the [-1,1] range.

Let's simulate some data with this features. For this example, we will fix gamma
to be 0.3

```{r dgp}

set.seed(12315)
N     <- 1000
Nc    <- 20
theta <- rep(rnorm(Nc, 0.3), N/Nc)
X     <- rnorm(N, theta)
```

The posterior probability in this case is

$$
P(\theta, \gamma | X) \approx P(X | \theta, \gamma)P(\theta|\gamma)P(\gamma)
$$

```{r posterior}
lupost <- function(pars) {
  
  sum(dnorm(X, pars[1:Nc], log = TRUE)) +
    sum(dnorm(pars[1:Nc], pars[Nc + 1], log=TRUE)) +
    dunif(pars[Nc + 1], -1, log=TRUE)
  
}
```

In this case, we will be using an adaptive MCMC transition kernel. The adaptation
will take place every 100 steps by computing the variance covariance matrix of
the last 500 steps, and will start from the 2,000 step

```{r adaptive}
library(fmcmc)
adaptk <- kernel_new(
  proposal = function(env) {
    
    if ((env$i > 2000) && !(env$i %% 100)) {
      S <<- cov(env$ans[(env$i - 500):(env$i - 1), , drop=FALSE])*2.38^2/(Nc + 1)
    }
    
    env$theta0 + mvtnorm::rmvnorm(1, mean = Mu, sigma = S)
    
  },
  S  = diag(Nc + 1),
  Mu = rep(0, Nc + 1)
)
```


```{r second-try-adaptive}
ans <- MCMC(
  rep(0, Nc + 1),
  lupost, nsteps = 10000, thin = 10,
  kernel = adaptk
)
```

```{r}
plot(cbind(summary(ans)$statistics[,"Mean"], Theta = c(unique(theta), 0.3)))
abline(a = 0, b = 1)
```
