---
title: "Workflow with fmcmc"
author: "George G. Vega Yon"
date: "April 19th, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow with fmcmc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.dim=c(9, 9), out.width=600, out.height=600)
```

# Motivating example

We start by loading the dataset that the `mcmc` package includes. We will use
the `logit` data set to obtain a posterior distribution of the model parameters
using the `MCMC` function.

```{r loading-data, include = TRUE}
library(fmcmc)
data(logit, package = "mcmc")
out <- glm(y ~ x1 + x2 + x3 + x4, data = logit, family = binomial, x = TRUE)
beta.init <- as.numeric(coefficients(out))
```

To be able to use the Metropolis-Hastings MCMC algorithm the function should be
(in principle) the log unnormalized posterior. The following block of code,
which was extracted from the `mcmc` package vignette "MCMC Package Example"
creates the function that we will be using

```{r log-unnorm-posterior}
lupost_factory <- function(x, y) function(beta) {
    eta <- as.numeric(x %*% beta)
    logp <- ifelse(eta < 0, eta - log1p(exp(eta)), - log1p(exp(- eta)))
    logq <- ifelse(eta < 0, - log1p(exp(eta)), - eta - log1p(exp(- eta)))
    logl <- sum(logp[y == 1]) + sum(logq[y == 0])
    return(logl - sum(beta^2) / 8)
}

lupost <- lupost_factory(out$x, out$y)
```

Let's give it a first try. In this case we will use the beta estimates from the
estimated GLM model as a starting point for the algorithm, and we will ask it to
sample 1e4 points from the posterior distribution (`nsteps`).

```{r estimation1.0}
# to get reproducible results
set.seed(42)
out <- MCMC(
  initial = beta.init,
  fun     = lupost,
  nsteps  = 1e3
)
```

Since the resulting object is of class `mcmc` (from the `coda` R package), we
can use all the functions included in `coda` for model diagnostics:

```{r post-estimation}
library(coda)
plot(out[,1:3])
```

So this chain has very poor mixing, so let's try again by using a smaller scale
for the normal kernel proposal moving it from 1 (the default value) to .2:

```{r estimation1.1}
# to get reproducible results
set.seed(42)
out <- MCMC(
  initial = beta.init,
  fun     = lupost,
  nsteps  = 1e3,
  kernel  = kernel_normal(scale = .2) 
)
```

The `kernel_normal`, which is the default kernel in the `MCMC` function, returns
an object of class `fmcmc_kernel`. In principle it consists on a list of two
functions that are used by the `MCMC` routine: `proposal`, the proposal kernel
function, and `logratio`, the function that returns the log of the
Metropolis-Hastings ratio. We will talk more about `fmcmc_kernel` objects later.
Now, let's look at the first three variables of our model:

```{r}
plot(out[,1:3])
```

Better. Now, ideally we should only be using observations from the stationary
distribution. Let's give it another try checking for convergence every 1,000
steps using the `convergence_geweke`:

```{r estimation1.2}
# to get reproducible results
set.seed(42)
out <- MCMC(
  initial = beta.init,
  fun     = lupost,
  nsteps  = 1e4,
  kernel  = kernel_normal(scale = .2),
  conv_checker = convergence_geweke(200)
)
```


A bit better. As announced by `MCMC`, the chain has reach a stationary state.
With this in hand, we can now rerun the algorithm such that we start from the
last couple of step of the chain, this time, without convergence monitoring as
it is no longer necessary.

We will increase the number of steps (sample size), use 2 chains using parallel
computing, and add some thinning to reduce autocorrelation:

```{r estimation2.0}
# Now we change the seed so we get a different stream of
# pseudo random numbers
set.seed(112) 

out_final <- MCMC(
  initial   = out,                       # Automagically takes the last 2 points
  fun       = lupost, 
  nsteps    = 5e4,                       # Increasing the sample size
  kernel    = kernel_normal(scale = .2),
  thin      = 10,
  nchains   = 2L,                        # Running parallel chains
  multicore = TRUE                       # in parallel.
  )
```

Observed that, instead of specifying what are the 2 starting points for each
chain, we passed the `out` to the initial set of parameters. By default, if
`initial` is of class `mcmc`, `MCMC` will take the last `nchains` points from
the chain as starting point for the new sequence. If `initial` is of class
`mcmc.list`, the number of chains in `initial` must match the `nchains`
parameter. We now see that the output posterior distribution appears to be
stationary

```{r final-results}
plot(out_final[, 1:3])
summary(out_final[, 1:3])
```


# Reusing fmcmc_kernel objects

`fmcmc_kernel` objects are enviroments that are passed to the `MCMC` function.
In the case of adaptive algorithms, users may want to reuse the adaptation.
